The unequal distribution of access to education and transport: insights from the São Paulo Metropolitan Region

### 2017-04-02 paper_v2.tex
Updated paper

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Robin Lovelace (University of Leeds) and Frederico R. Ramos (Fundação Getulio Vargas)},
            pdftitle={The unequal distribution of access to education and transport: insights from the São Paulo Metropolitan Region},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{The unequal distribution of access to education and transport: insights
from the São Paulo Metropolitan Region}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Ana I. Moreno-Monroy (Autonomous university of Barcelona), Robin Lovelace
(University of Leeds) and Frederico R. Ramos (Fundação Getulio Vargas)}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \date{}
  \predate{}\postdate{}

@@

\maketitle

\section{Abstract}\label{abstract}

In many large Latin American urban areas such as the São Paulo
Metropolitan Region (SPMR), growing social and economic inequalities are
embedded through high spatial inequality in the provision of education
and transport systems. This paper sheds light on the transport-education
inequality nexus with reference to a new school accessibility measure
applied to the SPMR. Our index combines information on the spatial
distribution of adolescents, the location of existing schools, and the
travel infrastructure serving the school catchment area into a single
measure. The index is used to measure school accessibility locally
across 633 areas within the SPMR. We use the index to simulate the
impact of a policiy aiming at increasing the centralization of public
secondary education provision, and find that it negatively affects
students with the lowest levels of accessibility. These results
illustrate how existing inequalities are amplified by variable
accessibility to schools across income groups and geographical space.

\bigskip

\textit{This version: January 2017. Preliminary version. Please do not cite}

\section{Introduction}\label{introduction}

Inequalities in educational and transport infrastructure are mutually
reinforcing: the right to mobility is intrinsically linked to the right
to education. Travel to school options are vital for ensuring a more
equitable supply of educational opportunity to diverse groups.
Conversely, poor accessibility to and from deprived areas can reinforce
social inequalities, with long-term implications. Against this
background, in this paper we propose a new way to measure school
accessibility in local areas within an urban area, and apply it for the
case of the São Paulo Metropolitan Region (SPMR).

@@

For this work we used the Google Matrix Distance API
(API stands for `application programming interface' and is used for
compter-computer communication over the web), a multi-modal
real-time travel data provider (see https://developers.google.com/maps/documentation/distance-matrix/). 
The API was used in preference to other options,
such as Routino, described in a paper by Singleton (2014),
and the OpenStreetMap Routing Machine (OSRM)
because its
ease of use (requiring no new software to be installed locally)
and performance: the Google Distance Matrix
API returns
route distance, price and times for journeys between origins and
destinations, provided either as Longitude/Latidue pairs or
as text strings to be 'geocoded' (converted to geographical location).

@@

Our work contributes to the literature on transport-related social
exclusion (Jones and Lucas 2012) (Lucas et al. 2016) by providing a
quantitative way to assess multiple dimensions into a single measure.
Education and public transport are both essential public services, and
thus are part of the set of fundamental and universal human rights. Lack
of sufficient access to these services limits the way in which
individuals use can their capacities and exersise their rights in a
context of equal opportunities (Gomide 2003). Our index provides a
benchmark to compare access inequalities in countries with unequal
provision of public services (Gomide 2006), (Pacione 1989). Considering
contestation as the process through which social groups mobilize in an
organized way in order to impede the implementation of unwanted
policies, or force the negotiation of new conditions, our index allows
embedded inequalities to be better understood.

@@

\section{2. Data}\label{data}

The school accessibility index was created using data from three main
sources: a 2008 School Census; school location data; and geographical
route data generated by the Google Matrix API.

The School Census data was provided by the
Brazilian National Institute of Educational Studies and Surveys (INEP).
The dataset contains information on all public educational
institutions, including a unique identifier and the number of students
enrolled in secondary education. School coordinates contained in a
database provided by the Center for Metropolitan Studies CEM
(\textit{Centro de Estudos da Metrópole}) for 2001 were used to
geolocate the schools in 2008 using the unique identifier.
To include new
schools built between 2001 and 2008, we used a dataset provided by the
Fundação SEADE (São Paulo State Agency for Statistical Analysis)
containing the goecoded location for all state administration schools at
2008. The this dataset set was obtained merging this two data source
using the standard ID used by the INEP. We geo-localized a total of
4,612 public schools in the SPRM.

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(stplanr)}
\NormalTok{o =}\StringTok{ }\KeywordTok{c}\NormalTok{(-}\FloatTok{46.31813}\NormalTok{, -}\FloatTok{23.38584}\NormalTok{) }\CommentTok{# coordinates of origin}
\NormalTok{d =}\StringTok{ }\KeywordTok{c}\NormalTok{(-}\FloatTok{46.34078}\NormalTok{, -}\FloatTok{23.37746}\NormalTok{) }\CommentTok{# coordinates of destination}
\KeywordTok{dist_google}\NormalTok{(}\DataTypeTok{from =} \NormalTok{o, }\DataTypeTok{to =} \NormalTok{d, }\DataTypeTok{mode =} \StringTok{"transit"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   distances duration
## 1      2962     2118
\end{verbatim}

The code above takes an origin (\texttt{o}) and destination (\texttt{d})
and finds the travel time and distance. This is reported as 2,962 metres
and 2,118 seconds (35 minutes) respectively. To ensure that the result
was relevant to school travel, the times were calculated for arrival
time at 9am on a weekday. The function was repeated for all
origin-destination pairs. To ensure the correct OD data was being
generated, the results were plotted on an interactive map (Figure 2).
The results show the high variability of route distance versus Euclidean
distance (`circuity'), providing further evidence of inequalities in
public transport provision in the city (Figure 3).
\begin{figure}
\includegraphics[width=1\linewidth]{od-map} \caption{Visual representation of the 12,697 OD pairs routed through the Google Distance Matrix API on an interactive map in RStudio, an open source data analysis platform.}\label{fig:unnamed-chunk-4}
\end{figure}
\begin{figure}
\includegraphics[width=1\linewidth]{public-vs-euclid-dists} \caption{Euclidean (straight line) vs public transport distances resulting from routing the 12,697 OD pairs of interest through the Google Distance Matrix API. The colours represent the interquantile range of the Euclidean distances travelled to school.}\label{fig:euclid}
\end{figure}

\section{3. Area of study}\label{area-of-study}

The São Paulo Metropolitan Region (SPMR) is a large metropolis located
in the South West of Brazil. It extends for over 7,700 squared
kilometers and groups 38 different municipalities besides the
municipality of São Paulo itself. The estimated population in 2010 was
close to 19.5 million in 2010. The integrated transport system, to which
users can access through an electronic card (Bilhete Único) is composed
of a railway and bus network. The railway network comprises 78.4
kilometers in five subway lines and six suburban railway lines which
provide less frequent and slower service than the subway. Some of the
bus network operates on dedicated bus corridors, which amounted to 500
km in 2015. As can be seen in Figure 1, the richer areas of the SPMR are
better served by public transport, while an important number of lower
income areas is under-served.

\begin{figure}
\includegraphics[width=0.8\linewidth]{income_pc} \caption{Public transport system and income per capita in São Paulo, 2010.}\label{fig:inc-trans}
\end{figure}

Regarding the educational system, since the constitution of 1946,
primary education is compulsory. In 2005, the legislation changed to
increase the duration of the primary education, going from eight to nine
years. In 2010, all regions finished the transition. In 2009, a
constitutional amendment imposed compulsory education for the primary
and secondary levels. The constitution imposes that this resolution
should be fully implemented by 2016. By law, the state/municipalities
should guaranty the adequate provision (100\% coverage of the demand).
There are national and state educational funds that are distributed
among the schools according to the number of enrolled students and
existence of facilities (library, laboratories, etc.).

The structure of the educational system is composed by both public and
private schools. The public schools are divided in the following levels:
Pre school (\textit{clase de alfabetização}) -- 4-5 year-olds (municipal
administration); Fundamental (\textit{ensino fundamental}) comprises
nine years, and the age range is usually 6-14 year-olds (municipal
administration); Secondary (\textit{ensino médio}) lasts a minimum of
three years, and the age range is 15-18 year-olds (state
administration); and Post-secondary (\textit{ensino superior}) (state
and federal administration).

@@

As can be seen by comparing Figures 4 and 5, with Figure 1, students in
the SPMR disproportionally attend private secondary schools in higher
income areas, while the opposite is true for public secondary schools.
\begin{table}[]\centering \caption{Modal split by type of secondary school}
\begin{tabular}{lcc} \hline
\multicolumn{3}{c}{\textbf{Secondary Public School Students}}  \\  \hline
 & Percent & Duration (min)  \\ \hline
Public & 23 & 36.7 \\
Private & 9 & 15.8 \\
Active & 68 & 15.7 \\ \hline
\multicolumn{3}{c}{\textbf{Secondary Private School Students}}  \\  \hline
Public & 28 & 35.7 \\ 
Private & 47 & 17.3 \\
Active & 25 & 13.9 \\\hline
\multicolumn{3}{l}{\small Public transport includes bus, metro and railway.} \\
\multicolumn{3}{l}{\small Active transport includes walking and cycling.}  
\end{tabular}
\end{table}
\begin{figure}
\includegraphics[width=0.7\linewidth]{med_pu} \caption{The home location of students attending public secondary schools students in the SPMR.}\label{fig:unnamed-chunk-5}
\end{figure}\begin{figure}
\includegraphics[width=0.7\linewidth]{med_pr} \caption{The home location of students attending private secondary schools students in the SPMR.}\label{fig:unnamed-chunk-6}
\end{figure}

As Figure 6 shows, enrolment in public secondary schools is more
dispersed spatially, as in principle public schools are not clearly
geographically concentrated within the region. There is, however, an
important degree of variation in the school seats in each sub-area,
ranging from zero to 7,754.
\begin{figure}
\includegraphics[width=0.7\linewidth]{enrol} \caption{Enrollment in public secondary schools in the SPMR.}\label{fig:unnamed-chunk-7}
\end{figure}
Table 1 shows the model split by type of secondary school
student.\footnote{The source of this information is the 2007
  Origin-Destination Household Travel Survey (O-D Survey), carried out
  by the São Paulo Metropolitan Transport Authority Mêtro. We focus on
  trips made by secondary school students with an indicated motive
  ``education'' at the destination. The total number of trips (excluding
  missing values) was 11,845. Of these, 5.5 percent were multimodal. In
  those cases, we assign the mode of the trip leg with the largest
  duration. We add all trip leg durations (in minutes) to obtain the
  total trip duration.} Clearly, secondary school students attending
private schools disproportionally travel by car, compared to public
school students (9\% versus 47\%), even though higher income areas are
better served by public transport. This is in line with the findings of
S{á} et al. (2015). Around 67\% of public school students commute to
school by active travel (mostly walking, since the percentage of biking
is still relatively small), while 23\% commute by public transport,
which has much longer mean duration than other modes. Gomide (2003),
using data from the 1997 SPMR Origin-Destination survey, reports that
the motive ``education'' explained at least 60\% of trips in the SPMR of
individuals earning up to one minimum wage a month. This proportion
decreases as income increases, as other motives such as work and
leisure become more relevant.

@@

\section{4. Method: measuring school
accessibility}\label{method-measuring-school-accessibility}

@@

\[ f(C_{ij}) = \left\{ 
                \begin{array}{ll}
                  1\quad if\quad C_{ij}<=t\\
                  0\quad if\quad C_{ij}<t
                \end{array}
              \right.
\]

@@

We calculate both measures for commutes to school by public transport.
We set a threshold of 30 minutes based on the mean average travel times
by this mode (see Table 1). We assign a minimum time of zero for
commutes within the same area (i.e., schools within the same area are
added to the index in all cases).

\section{5. Results}\label{results}

Figure 7 shows the spatial distribution of the cumulative-opportunity
measure. The index has a mean value of 3,777, which indicates that for
teens residing in the average area, there are 3,777 school seats that
can be reached within a journey by public transport of up to 30 minutes.
The standard deviation is 3,485 indicates large variation in the level
of accessibility across areas.

As can be seen on the Figure 7, areas with low accessibility include not
only the expanded financial business district located on the central
south west area, where higher-income teens mostly attend private school
(see Figures 1 and 5), but also areas with deficient public transport
provision, such as some north and south eastern areas.

\begin{figure}
\includegraphics[width=0.7\linewidth]{cum_index} \caption{Cumulative opportunity index}\label{fig:unnamed-chunk-8}
\end{figure}

Figure 8 shows the spatial distribution of the competitive accessibility
measure. Although it largely follows the distribution of the cumulative
index, it corrects the value in some areas based on the place of
residence of students across the SPRM. To interpret the value of the
index, recall that if all teens in an area attended secondary schools in
public secondary schools in their area (i.e., no student would attend
private school or travel outside her own area to attend school), the
index would take the value of 1. Then, in areas with a value lower than
one, the provision of public schools and public transport is such that
there is local under-provision, in the sense that the number of schools
seats that can be reached from the area is smaller than the number of
teens that can potentially reach the area (the opposite is true for
values higher than one). Note that some relatively close by areas,
especially in the eastern part of the city, have striking differences in
the level of accessibility. In some cases, an area can have twice the
level of accessibility than the most proximate area. This can be due to
differences in the local provision of public schools, public transport
or the number of students living within the reach of the areas.
\begin{figure}
\includegraphics[width=0.7\linewidth]{comp_index} \caption{Competitive accessibility index}\label{fig:unnamed-chunk-9}
\end{figure}
To understand the effects of changes in public school provision on
school accessibility, we conduct a comparative static analysis using our
competitive index. We compare the changes in school-accessibility by
accessibility quantiles before and after the implementation of a policy
that reallocates students attending school in areas with lower than
average provision to schools in areas with higher than average
provision. The experiment mimics a policy that aims at increasing scale
economies in the provision of public education by centralizing provision
in larger educational establishments, which is the rationale behind the
policy proposal made by the São Paulo state government (Ortellado 2015).

To implement the policy experiment, we assign students in areas with
lower than average enrolment to the next area with higher than average
enrolment, while keeping the distribution of teens and public transport
constant. In other words, total enrolment is the same before and after
the change, but the distribution of public secondary schools increases
in areas where it was already higher than average, and disappears in
areas where it was lower than average. The change is equivalent to
closing down 1,691 schools, or reallocating 163,231 students. Although
seemingly extreme, it is worth noting again that the proposal of the São
Paulo state government was estimated to impact 300,000 secondary
students attending public school (Ortellado 2015). We then compare the
changes in school accessibility for different quartiles before and after
this change. Table 2 summarizes the results.

\begin{longtable}[c]{@{}lrrr@{}}
\caption{Mean accessibility before and after
redistribution}\tabularnewline
\toprule
Quantile & Pre-change & Post-change & Change (\%)\tabularnewline
\midrule
\endfirsthead
\toprule
Quantile & Pre-change & Post-change & Change (\%)\tabularnewline
\midrule
\endhead
1st & 0.446 & 0.350 & -21.566\tabularnewline
2nd & 0.844 & 0.822 & -2.592\tabularnewline
3rd & 1.843 & 2.124 & 15.213\tabularnewline
4th & 4.280 & 5.483 & 28.099\tabularnewline
\bottomrule
\end{longtable}

\newpage

>>>>

\documentclass[3p,authoryear,preprint,review,12pt]{elsarticle} 
\usepackage{lipsum}
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{}%
 \let\@evenfoot\@oddfoot}
\makeatother
\usepackage{natbib,amssymb,amsmath,latexsym,makeidx,lscape,epic,overpic,graphicx,graphics,bm,pdflscape,rotating,multirow,bigdelim,color,epsfig,pgfpages} \usepackage{mathpazo}
\usepackage{mathtools}
\usepackage[portuguese, english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[left]{eurosym}
\usepackage{pdfsync,epstopdf,dblfloatfix,multicol,wrapfig,multirow}
\usepackage{epstopdf}
\usepackage{units}
\epstopdfsetup{outdir=./}
\usepackage[draft]{hyperref} % Drop [draft] to have references hyperlinked.
\usepackage{float}
\usepackage{longtable,booktabs}
\usepackage[toc,page]{appendix}
\usepackage{url}

\newcommand{\squeezeup}{\vspace{-2pt}}
\setlength{\textfloatsep}{3pt plus 1.0pt minus 2.0pt}
       \setlength{\abovecaptionskip }{-1ex}

%Standard settings for an Elsevier article. 
\def\elsartstyle{\def\normalsize{\@setfontsize\normalsize\@xiipt{12.5}} \def\small{\@setfontsize\small\@xipt{13.6}} \let\footnotesize=\small
   \def\large{\@setfontsize\large\@xivpt{18}} \def\Large{\@setfontsize\Large\@xviipt{22}}  \skip\@mpfootins = 18\p@ \@plus 2\p@ \normalsize }

@@

\begin{frontmatter}
 \title{The unequal distribution of access to education and transport: insights from the São Paulo Metropolitan Region}

  \author[uab]{Ana~I.~Moreno-Monroy}
 \ead{anaisabel.moreno@uab.cat}

  \author[uol]{Robin Lovelace}

   \author[fgv]{Frederico R. Ramos}

 \address[urv]{Applied Economics Department, Autonomous University of Barcelona}
  \address[uol]{Intitute of Trasnport Studies, University of Leeds}
 \address[fgv]{CEPESP, Fundação Getulio Vargas}

\begin{abstract}
In many large Latin American urban areas such as the São Paulo
Metropolitan Region (SPMR), growing social and economic inequalities are embedded through high spatial inequality in the provision of education and transport systems. This paper sheds light on the transport-education inequality nexus with reference to a new school accessibility measure applied to the SPMR. Our index combines information on the spatial distribution of adolescents, the location of existing schools, and the travel infrastructure serving the school catchment area into a single measure. The index is used to measure school accessibility locally across 633 areas within the SPMR. We use the index to simulate the impact of a policy aiming at increasing the centralization of public secondary education provision, and find that it negatively affects students with the lowest levels of accessibility. These results illustrate how existing inequalities are amplified by variable accessibility to schools across income groups and geographical space.
\end{abstract}

\begin{keyword}
Public transport, accessibility, schools, inequality, Latin America\\
\end{keyword}

\end{frontmatter}

\newpage
\section{Introduction}\label{introduction}

Inequalities in educational and transport infrastructure are mutually reinforcing: the right to mobility is intrinsically linked to the right to education. Travel to school options are vital for ensuring a more equitable supply of educational opportunity to diverse groups.Conversely, poor accessibility to and from deprived areas can reinforce social inequalities, with long-term implications. Against this background, in this paper we propose a new way to measure school accessibility in local areas within an urban area, and apply it for the case of the São Paulo Metropolitan Region (SPMR).

@@

For this work we used the Google Matrix Distance API
(API stands for `application programming interface' and is used for
computer-computer communication over the web), a multi-modal
real-time travel data provider (see \url{https://developers.google.com/maps/documentation/distance-matrix/}). 
The API was used in preference to other options,
such as Routino, described in a paper by Singleton (2014),
and the OpenStreetMap Routing Machine (OSRM)
because its
ease of use (requiring no new software to be installed locally)
and performance: the Google Distance Matrix
API returns
route distance, price and times for journeys between origins and
destinations, provided either as Longitude/Latitude pairs or
as text strings to be 'geocoded' (converted to geographical location).

@@

Our work contributes to the literature on transport-related social
exclusion (Jones and Lucas 2012) (Lucas et al. 2016) by providing a
quantitative way to assess multiple dimensions into a single measure. Education and public transport are both essential public services, and thus are part of the set of fundamental and universal human rights. Lack of sufficient access to these services limits the way in which individuals use can their capacities and exercise their rights in a context of equal opportunities (Gomide 2003). Our index provides a benchmark to compare access inequalities in countries with unequal provision of public services (Gomide 2006), (Pacione 1989). Considering contestation as the process through which social groups mobilize in an organized way in order to impede the implementation of unwanted policies, or force the negotiation of new conditions, our index allows embedded inequalities to be better understood.

@@

\section{Data}\label{data}

The school accessibility index was created using data from the following sources: 
\begin{itemize}
\item Demographic Census – 2010 – IBGE
\item Origen – Destination Survey – 2007 – Metro
\item School Census – 2008 – INEP
\item Geocoded Schools – CEM/CEBRAP – 2001
\item Geocoded Public Schools – SEADE – 2008
\item Google Distance Matrix API 
\end{itemize}

The School Census data was provided by the Brazilian National Institute of Educational Studies and Surveys (INEP).
The dataset contains information on all public educational
institutions, including a unique identifier and the number of students enrolled in secondary education. School coordinates contained in a database provided by the Center for Metropolitan Studies CEM (\textit{Centro de Estudos da Metrópole}) for 2001 were used to geolocate the schools in 2008 using the unique identifier.
To include new schools built between 2001 and 2008, we used a dataset provided by the Fundação SEADE (São Paulo State Agency for Statistical Analysis) containing the geocoded location for all state administration schools at 2008. The this dataset set was obtained merging this two data source using the standard ID used by the INEP. We geo-localized a total of 4,612 public schools in the SPMR.

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{box} \label{fig:box}
\end{figure}
\begin{verbatim}
##   distances duration
## 1      2962     2118
\end{verbatim}

The code above takes an origin (\texttt{o}) and destination (\texttt{d}) and finds the travel time and distance. This is reported as 2,962 metres and 2,118 seconds (35 minutes) respectively. To ensure that the result was relevant to school travel, the times were calculated for arrival time at 9am on a weekday. The function was repeated for all origin-destination pairs. To ensure the correct OD data was being generated, the results were plotted on an interactive map (Figure \ref{fig:pairs}).
The results show the high variability of route distance versus Euclidean distance (`circuity'), providing further evidence of inequalities in public transport provision in the city (Figure \ref{fig:euclid}).
\begin{figure}[H]
\includegraphics[width=1\linewidth]{od-map} \caption{Visual representation of the 12,697 OD pairs routed through the Google Distance Matrix API on an interactive map in RStudio, an open source data analysis platform.}\label{fig:pairs}
\end{figure}

\begin{figure}[H]
\includegraphics[width=0.5\linewidth]{public-vs-euclid-dists} \caption{Euclidean (straight line) vs public transport distances resulting from routing the 12,697 OD pairs of interest through the Google Distance Matrix API. The colours represent the interquantile range of the Euclidean distances travelled to school.}\label{fig:euclid}
\end{figure}

\section{Area of study}\label{area-of-study}

The São Paulo Metropolitan Region (SPMR) groups 38 different municipalities besides the municipality of São Paulo itself. The estimated population in 2010 was close to 19.5 million inhabitants. It has an integrated transport system, to which users can access through an electronic card (Bilhete Único) composed of a railway and bus network. The railway network comprises 78.4 kilometers in five subway lines and six suburban railway lines which provide less frequent and slower service than the subway. Some of the bus network operates on dedicated bus corridors, which amounted to 500 km in 2015. As can be seen in Figure \ref{fig:inc-trans}, the richer areas of the SPMR are better served by public transport, while an important number of lower income areas is under-served.

\begin{figure}[H]
\includegraphics[width=0.8\linewidth]{income_pc} \caption{Public transport system and income per capita in São Paulo, 2010.}\label{fig:inc-trans}
\end{figure}

Regarding the educational system, primary education is compulsory. In 2005, the legislation changed to increase the duration of the primary education, going from eight to nine years. In 2010, all regions finished the transition. In 2009, a constitutional amendment imposed compulsory education for the primary and secondary levels. The constitution imposes that this resolution should be fully implemented by 2016. By law, the state/municipalities should guaranty the adequate provision (100\% coverage of the demand). There are national and state educational funds that are distributed among the schools per number of enrolled students and existence of facilities (library, laboratories, etc.). The structure of the educational system is composed by both public and private schools. The public schools are divided in the following levels: Pre school (classe de alfabetização) – 4-5 year-olds (municipal administration); Fundamental (ensino fundamental) comprises nine years, and the age range is usually 6-14 year-olds (municipal administration); Secondary (ensino médio) lasts a minimum of three years, and the age range is 15-18 year-olds (state administration); and Post-secondary (ensino superior) (state and federal administration).

@@

As can be seen by comparing Figures \ref{fig:inc-trans} and \ref{fig:students}, students in the SPMR disproportionally attend private secondary schools in higher income areas, while the opposite is true for public secondary schools.

\begin{figure}[H]
\includegraphics[width=1\linewidth]{med_pu} \caption{The home location of students attending public secondary schools students in the SPMR.}\label{fig:students}
\end{figure}

As Figure 5 shows, enrolment in public secondary schools is more
dispersed spatially, as in principle public schools are not clearly
geographically concentrated within the region. There is, however, an
important degree of variation in the school seats in each sub-area,
ranging from zero to 7,754.
\begin{figure}
\includegraphics[width=0.7\linewidth]{enrol} \caption{Enrollment in public secondary schools in the SPMR.}\label{fig:unnamed-chunk-7}
\end{figure}
Table 1 shows the modal split by type of secondary school student.\footnote{The source of this information is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the São Paulo Metropolitan Transport Authority Mêtro. We focus on   trips made by secondary school students with an indicated motive   ``education'' at the destination. The total number of trips (excluding missing values) was 11,845. Of these, 5.5 percent were multimodal. In those cases, we assign the mode of the trip leg with the largest duration. We add all trip leg durations (in minutes) to obtain the total trip duration.}

\begin{table}[H]\centering \caption{Modal split by type of secondary school}
\begin{tabular}{lcc} \hline
\multicolumn{3}{c}{\textbf{Secondary Public School Students}}  \\  \hline
 & Percent & Duration (min)  \\ \hline
Public & 23 & 36.7 \\
Private & 9 & 15.8 \\
Active & 68 & 15.7 \\ \hline
\multicolumn{3}{c}{\textbf{Secondary Private School Students}}  \\  \hline
Public & 28 & 35.7 \\ 
Private & 47 & 17.3 \\
Active & 25 & 13.9 \\\hline
\multicolumn{3}{l}{\small Public transport includes bus, metro and railway.} \\
\multicolumn{3}{l}{\small Active transport includes walking and cycling.}  
\end{tabular}
\end{table} \label{split}

Clearly, secondary school students attending private schools disproportionally travel by car, compared to public school students (47\% versus 9\%), even though higher income areas are better served by public transport. This is in line with the findings of
S{á} et al. (2015). Around 67\% of public school students commute to
school by active travel (mostly walking, since the percentage of biking is still relatively small), while 23\% commute by public transport, which has much longer mean duration than other modes. Gomide (2003), using data from the 1997 SPMR Origin-Destination survey, reports that the motive ``education'' explained at least 60\% of trips in the SPMR of individuals earning up to one minimum wage a month. This proportion decreases as income increases, as other motives such as work and leisure become more relevant.

@@

\section{Method: measuring school
accessibility}\label{method-measuring-school-accessibility}

@@

\[ f(C_{ij}) = \left\{ 
                \begin{array}{ll}
                  1\quad if\quad C_{ij}<=t\\
                  0\quad if\quad C_{ij}>t
                \end{array}
              \right.
\]

@@

We calculate both measures for commutes to school by public transport. We set a threshold of 30 minutes based on the mean average travel times by this mode (see Table \ref{split}). As a robustness check, in the result sections we show results for 35 and 40 minutes thresholds. We assign a minimum time of zero for commutes within the same area (i.e., schools within the same area are added to the index in all cases).

\section{Results}\label{results}

\subsection{School accessibility}\label{results}

Figure \ref{fig:cum_ind} shows the spatial distribution of the cumulative-opportunity measure. The index has a mean value of 3,777, which indicates that for teens residing in the average area, there are 3,777 school seats that can be reached within a journey by public transport of up to 30 minutes. The standard deviation is 3,485 indicates large variation in the level of accessibility across areas.

The results show that areas with low accessibility include not
only the expanded financial business district located on the central
south west area, where higher-income teens mostly attend private school (see Figures \ref{fig:inc-trans} and \ref{fig:students}), but also areas with deficient public transport provision, such as some north and south eastern areas.

\begin{figure}[H]
\includegraphics[width=0.7\linewidth]{cum_index} \caption{Cumulative opportunity index}\label{fig:cum_ind}
\end{figure}

Figure \ref{fig:comp_ind} shows the spatial distribution of the competitive accessibility measure. Although it largely follows the distribution of the cumulative index, it corrects the value in some areas based on the place of residence of students across the SPMR. To interpret the value of the index, recall that if all teens in an area attended secondary schools in public secondary schools in their area (i.e., no student would attend private school or travel outside her own area to attend school), the index would take the value of 1. Then, in areas with a value lower than one, the provision of public schools and public transport is such that there is local under-provision, in the sense that the number of schools seats that can be reached from the area is smaller than the number of teens that can potentially reach the area (the opposite is true for values higher than one). Note that some relatively close by areas, especially in the eastern part of the city, have striking differences in the level of accessibility. In some cases, an area can have twice the level of accessibility than the most proximate area. This can be due to differences in the local provision of public schools, public transport or the number of students living within the reach of the areas.

\begin{figure}[H]
\includegraphics[width=0.7\linewidth]{comp_index} \caption{Competitive accessibility index}\label{fig:comp_ind}
\end{figure}

\subsection{Policy experiment}\label{results}

To understand the effects of changes in public school provision on
school accessibility, we conduct a comparative static analysis using our competitive index. The experiment mimics a policy that aims at increasing scale economies in the provision of public education by centralizing provision in larger educational establishments, which is the rationale behind the policy proposal made by the São Paulo state government (Ortellado 2015). In this spirit, instead of selecting a specific number of schools to be closed or students to be reallocated, we compare the changes in school-accessibility before and after the implementation of a policy that closes down schools which fall under a specified size threshold. The threshold  choice is based on the quantiles of the school size distribution of all schools in the area. Figure \ref{density} shows the density plot of the number of students and three different thresholds: 15\%, 25\% and 50\%.

\begin{figure}[H]
\includegraphics[width=0.7\linewidth]{density} \caption{Competitive accessibility index}\label{fig:density}
\end{figure}

Table \ref{thresholds} summarizes the size threshold corresponding to each scenario, the number of schools which would close and the number of students which would be reallocated in each case.

\begin{table}[H]\caption{Policy experiment scenarios}\label{thresholds}
\begin{tabular}[c]{lccc}\hline
Scenario & Size threshold & \# Schools closed & \# Students reallocated \\ \hline
15\% & 337 & 343 & 8,002 \\
25\% & 572 & 711 & 36,807 \\
50\% & 1,061 & 1691 & 163,231 \\\hline
\end{tabular}
\end{table}
